Face Recognition: Too Bias, or Not Too Bias?

### 2020-03-11 manuscript.tex
arXiv
added cvpr template and started filling in

 \documentclass[a4paper, 10pt, conference]{ieeeconf}      % Use this line for a4

@@

\usepackage{FG2019}
\usepackage{multirow}
\newcommand{\minitab}[2][c]{\begin{tabular}{#1}#2\end{tabular}}
\usepackage[acronym]{glossaries}
\usepackage{acronym}
\usepackage{bigstrut}
% symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{amssymb}
\usepackage{amsmath}% assumes amsmath package installed
\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
% \usepackage{hyperref}       % hyperlinks
\usepackage[colorlinks = true,
            linkcolor = blue,
            urlcolor  = blue,
            citecolor = blue,
            anchorcolor = blue]{hyperref}

@@

\graphicspath{{images/}} %Setting the graphicspath
\usepackage{bibunits}

@@

\newcommand{\com}[1]{\textcolor{red}{#1}}
\newcommand{\ie}{\textit{i}.\textit{e}., }
\newcommand{\eg}{\textit{e}.\textit{g}., }
\newcommand*{\etc}{etc.\@\xspace}

@@

\FGfinalcopy % *** Uncomment this line for the final submission
\let\vec\mathbf

@@

\title{\LARGE \bf
Face Recognition: Too Bias, or Not Too Bias?
}

% \title{\LARGE \bf\vspace{-10mm}
% Face Recognition: Too Bias, or Not Too Bias?
% }

%use this in case of several affiliations
\author{\parbox{16cm}{\centering
    {\large Joseph P Robinson$^1$, Gennady Livitz$^2$, Yann Henon$^2$, Can Qin$^1$,\\ Yun Fu$^1$, and Samson Timoner$^2$}\\
    {\normalsize
    \hspace{-.4in}$^{1}$Northeastern University\hspace{.7in} $^{2}$ISM Connect}}
    \thanks{}% <-this % stops a space
}

\begin{document}
%\begin{bibunit}

\def\angle{0}
\def\radius{3}
\def\cyclelist{{"orange","blue","red","green","magenta","cyan"}}
\newcount\cyclecount \cyclecount=-1
\newcount\ind \ind=-1

\ifFGfinal
\thispagestyle{empty}
\pagestyle{empty}
\else
\author{Anonymous FG 2019 submission\\ Paper ID \FGPaperID \\}
\pagestyle{plain}
\fi
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
We reveal critical insights into problems of bias in state-of-the-art \gls{fr} systems using a novel \gls{bfw} dataset: data balanced for gender and ethnic groups. We show variations in the optimal scoring threshold for face-pairs across different subgroups. Thus, the conventional approach of learning a  global threshold for all pairs resulting in performance gaps among subgroups. By learning subgroup-specific thresholds, we not only mitigate problems in performance gaps but also show a notable boost in the overall performance. Furthermore, we do a human evaluation to measure the bias in humans, which supports the hypothesis that such a bias exists in human perception. For the \gls{bfw} database, source code, and more, visit \href{https://github.com/visionjo/facerec-bias-bfw}{github.com/visionjo/facerec-bias-bfw}
\end{abstract}
\
\input{sections/introduction.tex}

@@

% \vspace{-3mm}
\bibliographystyle{ieee}
\balance
\bibliography{references}

% \putbib
% \end{bibunit}
\begin{bibunit}
\input{supplemental.tex}
\end{bibunit}

>>>>

%  \documentclass[a4paper, 10pt, conference]{ieeeconf}      % Use this line for a4
\documentclass[10pt,twocolumn,letterpaper]{article}

\usepackage{cvpr}

@@

% \usepackage{FG2019}
\usepackage{multirow}
\usepackage{graphicx}

\newcommand{\minitab}[2][c]{\begin{tabular}{#1}#2\end{tabular}}
\usepackage[acronym]{glossaries}
\usepackage{acronym}
\usepackage{bigstrut}
\usepackage{cvpr}
\usepackage{times}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
% symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{amssymb}
\usepackage{amsmath}% assumes amsmath package installed
\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
% % \usepackage{hyperref}       % hyperlinks
% \usepackage[colorlinks = true,
%             linkcolor = blue,
%             urlcolor  = blue,
%             citecolor = blue,
%             anchorcolor = blue]{hyperref}
\usepackage[pagebackref=true,breaklinks=true,letterpaper=true,colorlinks,bookmarks=false]{hyperref}

@@

% \graphicspath{{figures/}} %Setting the graphicspath
\usepackage{bibunits}

@@

\newcommand{\com}[1]{\textcolor{red}{#1}}
% \newcommand{\ie}{\textit{i}.\textit{e}., }
% \newcommand{\eg}{\textit{e}.\textit{g}., }
% \newcommand*{\etc}{etc.\@\xspace}

@@

% \FGfinalcopy % *** Uncomment this line for the final submission
\let\vec\mathbf

@@

% \title{\LARGE \bf
% Face Recognition: Too Bias, or Not Too Bias?
% }

% % \title{\LARGE \bf\vspace{-10mm}
% % Face Recognition: Too Bias, or Not Too Bias?
% % }

% %use this in case of several affiliations
% \author{\parbox{16cm}{\centering
%     {\large Joseph P Robinson$^1$, Gennady Livitz$^2$, Yann Henon$^2$, Can Qin$^1$,\\ Yun Fu$^1$, and Samson Timoner$^2$}\\
%     {\normalsize
%     \hspace{-.4in}$^{1}$Northeastern University\hspace{.7in} $^{2}$ISM Connect}}
%     \thanks{}% <-this % stops a space
% }

% \cvprfinalcopy % *** Uncomment this line for the final submission

\def\cvprPaperID{****} % *** Enter the CVPR Paper ID here
\def\httilde{\mbox{\tt\raisebox{-.5ex}{\symbol{126}}}}

% Pages are numbered in submission mode, and unnumbered in camera-ready
\ifcvprfinal\pagestyle{empty}\fi
\begin{document}

%%%%%%%%% TITLE
\title{Face Recognition: Too Bias, or Not Too Bias?}

\author{First Author\\
Institution1\\
Institution1 address\\
{\tt\small firstauthor@i1.org}
% For a paper whose authors are all at the same institution,
% omit the following lines up until the closing ``}''.
% Additional authors and addresses can be added with ``\and'',
% just like the second author.
% To save space, use either the email address or home page, not both
\and
Second Author\\
Institution2\\
First line of institution2 address\\
{\tt\small secondauthor@i2.org}
}

\maketitle

% \begin{document}
%\begin{bibunit}

% \def\angle{0}
% \def\radius{3}
% \def\cyclelist{{"orange","blue","red","green","magenta","cyan"}}
% \newcount\cyclecount \cyclecount=-1
% \newcount\ind \ind=-1

% \ifFGfinal
% \thispagestyle{empty}
% \pagestyle{empty}
% \else
% \author{Anonymous FG 2019 submission\\ Paper ID \FGPaperID \\}
% \pagestyle{plain}
% \fi
% \maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
We reveal critical insights into problems of bias in state-of-the-art \gls{fr} systems using a novel \gls{bfw} dataset: data balanced for gender and ethnic groups. We show variations in the optimal scoring threshold for face-pairs across different subgroups. Thus, the conventional approach of learning a  global threshold for all pairs resulting in performance gaps among subgroups. By learning subgroup-specific thresholds, we not only mitigate problems in performance gaps but also show a notable boost in the overall performance. Furthermore, we do a human evaluation to measure the bias in humans, which supports the hypothesis that such a bias exists in human perception. For the \gls{bfw} database, source code, and more, visit \href{https://github.com/visionjo/facerec-bias-bfw}{github.com/visionjo/facerec-bias-bfw}
\end{abstract}

\input{sections/introduction.tex}

@@

% % \vspace{-3mm}
% \bibliographystyle{ieee}
% \balance
% \bibliography{references}

{\small
\bibliographystyle{ieee_fullname}
\bibliography{references}
}

% \putbib
% \end{bibunit}
% \begin{bibunit}
% \input{supplemental.tex}
% \end{bibunit}

### 2020-03-11 supplemental.tex
added cvpr template and started filling in

%
\newpage
\onecolumn
% \begin{minipage*}[\linewidth]{width of the minipage}
\renewcommand{\thesection}{\alph{section}}
\glsresetall
\setcounter{section}{0}

\section{Other \gls{cnn} models.}~\label{app:sec:other:models}
Variations in optimal threshold exist across models (Fig.~\ref{fig:sdm-appendix-a}). Like in Fig.~\ref{fig:detcurves}, the \gls{det} curves for three \gls{cnn}-based models, each trained on VGG2 with softmax but with different backbones.\footnote{Used pre-trained models and public Github, \href{https://github.com/rcmalli/keras-vggface}{https://github.com/rcmalli/keras-vggface}} Notice similar trends across subgroups and models, which is consistent with  Sphereface as well (Fig.~\ref{fig:detcurves}). For example, the plots generated with Spherface and VggFace2 all have the \gls{wm} curve at the bottom (\ie best) and \gls{af} on top (\ie worst). 
% ArcFace (not shown) shown similar results. We would have to cite + will incorporate later .... a blank statement and sphereface is comparable ... up to you i do not care
% \begin{figure}[h!]
% % \vspace{-2mm}
%     \centering
%     \includegraphics[width=.8\linewidth, trim={0mm 0mm 0mm 0mm},clip]{figures/SDM.pdf}\\
%     \caption{\textbf{\gls{det} curves for different CNN models}. \gls{fnr} (\%) (vertical) vs \gls{fpr} (log-scale)  (horizontal) for VGG2~\cite{Cao18} models with different backbones (vgg16, Resnet50~\cite{he2016deep}, SEnet50~\cite{hu2018squeeze}). Lower is better. For each plot, \gls{wm} is the best performing curve, \gls{af} is the worst. The the ordering of the curves is roughly the same for each backbone.}\label{fig:sdm-appendix-a}
%     \vspace{-3mm}
% \end{figure}

\begin{figure}[h!]
\vspace{-1mm}
    \centering
    \begin{subfigure}[t]{.3\linewidth}
    \includegraphics[width=.75\linewidth]{figures/curve_vgg16_subgroups-crop.pdf}
    \caption{VGG16~\cite{simonyan2014very}}
 \end{subfigure}
    \begin{subfigure}[t]{.27\linewidth}
    \includegraphics[width=.8\linewidth]{figures/curve_resnet50_subgroups-crop.pdf}
    \caption{ResNet50~\cite{he2016deep}}
   \end{subfigure}
    \begin{subfigure}[t]{.27\linewidth}
    \includegraphics[width=.8\linewidth]{figures/curve_senet50_subgroups-crop.pdf}
    \caption{SENet~\cite{hu2018squeeze}}
    \end{subfigure}
    \caption{\textbf{\gls{det} curves for different CNN models}. \gls{fnr} (\%) (vertical) vs \gls{fpr} (log-scale)  (horizontal) for VGG2~\cite{Cao18} models with different backbones (VGG16, Resnet50, SEnet50). Lower is better. For each plot, \gls{wm} is the best performing curve, \gls{af} is the worst. The the ordering of the curves is roughly the same for each backbone.}\label{fig:sdm-appendix-a}
    \vspace{-1mm}
\end{figure}

Sample faces per subgroup of the \gls{bfw} dataset are shown in Fig.~\ref{fig:montage:app}.
\begin{figure}[h!]
\vspace{-2mm}
    \centering
    \includegraphics[width=.7\linewidth]{figures/facemontage.pdf}
    \caption{\textbf{Sample of \gls{bfw}}. Each row depicts a different gender, \gls{f} (top) and \gls{m} (bottom). Columns are grouped by ethnicity (\ie \gls{a}, \gls{b}, \gls{i}, and \gls{w}, respectfully).}
    \label{fig:montage:app}
    \vspace{-1mm}
\end{figure}
{
\scriptsize
\begin{thebibliography}{widest entry}
 \bibitem[1]{Cao18} Qiong Cao, Li Shen, Weidi Xie, Omkar M. Parkhi, and Andrew Zisserman. ``Vggface2: A dataset for recognising faces across pose and age.'' In \textit{IEEE International Conference on Automatic Face \& Gesture Recognition.} 2018.
 \bibitem[2]{simonyan2014very} Karen Simonyan and Andrew Zisserman. ``Very deep convolutional networks for large-scale image recognition'' \textit{arXiv preprint arXiv:1409.1556.} 2014.
  \bibitem[3]{he2016deep} He, Kaiming, Xiangyu Zhang, Shaoqing Ren, and Jian Sun. ``Deep residual learning for image recognition.'' In \textit{IEEE Conference on Computer Vision and Pattern Recognition.} 2016.
 \bibitem[4]{hu2018squeeze} Hu, Jie, Li Shen, and Gang Sun. ``Squeeze-and-excitation networks.'' In \textit{IEEE Conference on Computer Vision and Pattern Recognition.} 2018.
\end{thebibliography}
}

>>>>

%
\newpage
\onecolumn

% \begin{minipage*}[\linewidth]{width of the minipage}
% \renewcommand{\thesection}{\alph{section}}
\glsresetall
% \setcounter{section}{0}
\begin{comment}
\section{Other \gls{cnn} models.}~\label{app:sec:other:models}
Variations in optimal threshold exist across models (Fig.~\ref{fig:sdm-appendix-a}). Like in Fig.~\ref{fig:detcurves}, the \gls{det} curves for three \gls{cnn}-based models, each trained on VGG2 with softmax but with different backbones.\footnote{Used pre-trained models and public Github, \href{https://github.com/rcmalli/keras-vggface}{https://github.com/rcmalli/keras-vggface}} Notice similar trends across subgroups and models, which is consistent with  Sphereface as well (Fig.~\ref{fig:detcurves}). For example, the plots generated with Spherface and VggFace2 all have the \gls{wm} curve at the bottom (\ie best) and \gls{af} on top (\ie worst). 
% ArcFace (not shown) shown similar results. We would have to cite + will incorporate later .... a blank statement and sphereface is comparable ... up to you i do not care
% \begin{figure}[h!]
% % \vspace{-2mm}
%     \centering
%     \includegraphics[width=.8\linewidth, trim={0mm 0mm 0mm 0mm},clip]{figures/SDM.pdf}\\
%     \caption{\textbf{\gls{det} curves for different CNN models}. \gls{fnr} (\%) (vertical) vs \gls{fpr} (log-scale)  (horizontal) for VGG2~\cite{Cao18} models with different backbones (vgg16, Resnet50~\cite{he2016deep}, SEnet50~\cite{hu2018squeeze}). Lower is better. For each plot, \gls{wm} is the best performing curve, \gls{af} is the worst. The the ordering of the curves is roughly the same for each backbone.}\label{fig:sdm-appendix-a}
%     \vspace{-3mm}
% \end{figure}

\begin{figure}[h!]
    \centering
    \begin{subfigure}[t]{.3\linewidth}
    \includegraphics[width=.75\linewidth]{figures/curve_vgg16_subgroups-crop.pdf}
    \caption{VGG16~\cite{simonyan2014very}}
 \end{subfigure}
    \begin{subfigure}[t]{.27\linewidth}
    \includegraphics[width=.8\linewidth]{figures/curve_resnet50_subgroups-crop.pdf}
    \caption{ResNet50~\cite{he2016deep}}
   \end{subfigure}
    \begin{subfigure}[t]{.27\linewidth}
    \includegraphics[width=.8\linewidth]{figures/curve_senet50_subgroups-crop.pdf}
    \caption{SENet~\cite{hu2018squeeze}}
    \end{subfigure}
    \caption{\textbf{\gls{det} curves for different CNN models}. \gls{fnr} (\%) (vertical) vs \gls{fpr} (log-scale)  (horizontal) for VGG2~\cite{Cao18} models with different backbones (VGG16, Resnet50, SEnet50). Lower is better. For each plot, \gls{wm} is the best performing curve, \gls{af} is the worst. The the ordering of the curves is roughly the same for each backbone.}\label{fig:sdm-appendix-a}
\end{figure}
\end{comment}

\begin{table}
\begin{center}
\glsunsetall
\caption{Intended \gls{fpr} per subgroup. Each tuple represents percentage the respective rate was achieved (\textcolor{green}{+\%} or \textcolor{red}{-\%}) paired with FNR. Each subgroup shows the use of $t_g$ (top) and individual $t_o$ (bottom).}
\begin{tabular}{l c c c c c}
     \gls{fpr} & 0.3 & 0.1 & 0.01 & 0.001 & 0.0001\\\midrule
    \multirow{2}{.1mm}{\textbf{\gls{af}}} & (\textcolor{red}{-4.5}, )&(\textcolor{red}{-13.4}, )  &(\textcolor{red}{-25.8}, ) &(\textcolor{red}{-25.8}, )  &(\textcolor{red}{-25.5}, )  \\[-4pt]
        & (0.0, ) &(0.0, )  & (0.0, ) &  (\textcolor{green}{0.1}, )&(\textcolor{red}{-5.6}, )  \\[-1pt]

    \multirow{2}{3mm}{\textbf{\gls{am}}}&(\textcolor{green}{2.3}, )&(\textcolor{green}{4.1}, )&(\textcolor{green}{11.2}, )&(\textcolor{green}{14.5}, )&(\textcolor{green}{52.1}, ) \\[-4pt]
        & (0.0, ) &(0.0, )  & (\textcolor{green}{0.2}, ) & (\textcolor{green}{0.2}, ) & (\textcolor{red}{5.4}, )\\[-1pt]

    \multirow{2}{3mm}{\textbf{\gls{bf}}}& (\textcolor{red}{-1.6}, ) &(\textcolor{red}{-8.2}, ) &(\textcolor{red}{-23.4}, ) &(\textcolor{red}{-27.7}, ) &(\textcolor{red}{-41.8}, ) \\[-4pt]
        & (0.0, ) &(\textcolor{red}{-0.1}, )  & (0.0, ) & (0.0, ) & (\textcolor{red}{5.7}, )\\[-1pt]

    \multirow{2}{3mm}{\textbf{\gls{bm}}}&(\textcolor{green}{2.3}, )&(\textcolor{green}{6.5}, )&(\textcolor{green}{17.6}, )&(\textcolor{green}{29.8}, )&(\textcolor{green}{95.8}, ) \\[-4pt]
        & (0.0, ) &(\textcolor{red}{-0.1}, )  & (0.0, ) & (\textcolor{green}{0.1}, ) & (\textcolor{red}{5.5}, )\\[-1pt]

    \multirow{2}{3mm}{\textbf{\gls{if}}} &(\textcolor{red}{-9.0}, ) &(\textcolor{red}{-26.1}, ) &(\textcolor{red}{-58.6}, ) &(\textcolor{red}{-46.1}, ) &(\textcolor{red}{-60.9}, )\\[-4pt]
        & (0.0, ) &(0.0, )  & (0.0, ) & (\textcolor{green}{0.2}, ) & (\textcolor{red}{5.4}, )\\[-1pt]

    \multirow{2}{3mm}{\textbf{\gls{im}}} &(\textcolor{green}{2.4}, )&(\textcolor{green}{4.3}, )&(\textcolor{red}{-3.4}, )&(\textcolor{red}{-16.4}, )&(\textcolor{red}{-76.2}, )\\[-4pt]
        & (0.0, ) &(0.0, )  & (\textcolor{green}{0.1}, ) & (\textcolor{green}{0.1}, )  &(\textcolor{red}{5.5}, ) \\[-1pt]

    \multirow{2}{3mm}{\textbf{\gls{wf}}} &(\textcolor{red}{-1.1},) &(\textcolor{green}{5.8}, ) &(\textcolor{green}{50.6}, ) &(\textcolor{green}{52.6}, ) &(\textcolor{green}{158.0}, )\\[-4pt]
       & (0.0, ) &(\textcolor{red}{-0.1}, )  & (\textcolor{green}{0.3}, ) & (\textcolor{green}{0.2}, )  &(\textcolor{red}{5.5}, )  \\[-1pt]

    \multirow{2}{3mm}{\textbf{\gls{wm}}} & (\textcolor{green}{10.5}, ) & (\textcolor{green}{38.6}, )& (\textcolor{green}{120.0}, )& (\textcolor{green}{83.4}, )& (\textcolor{green}{200.0}, ) \\[-4pt]

        & (0.0, ) &(\textcolor{green}{0.2}, ) &(\textcolor{green}{0.2}, ) & (\textcolor{green}{0.2}, )&(\textcolor{red}{-5.5}, )\\[-1pt]
    % \midrule
    % \multirow{2}{3mm}{\textbf{Avg.}} \\[-4pt]
    %   & (0.0, ) &(0.0, )  & (\textcolor{green}{-}, ) &  \\[-10pt]
\end{tabular}
\end{center}
\end{table}

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/SDM-crop.pdf}
    \caption{Caption}
    \label{fig:my_label}
\end{figure}

Sample faces per subgroup of the \gls{bfw} dataset are shown in Fig.~\ref{fig:montage:app}.
\begin{figure}[h!]
    \centering
    \includegraphics[width=.7\linewidth]{figures/facemontage.pdf}
    \caption{\textbf{Sample of \gls{bfw}}. Each row depicts a different gender, \gls{f} (top) and \gls{m} (bottom). Columns are grouped by ethnicity (\ie \gls{a}, \gls{b}, \gls{i}, and \gls{w}, respectfully).}
    \label{fig:montage:app}
\end{figure}
\begin{comment}
\scriptsize
\begin{thebibliography}{widest entry}
 \bibitem[1]{Cao18} Qiong Cao, Li Shen, Weidi Xie, Omkar M. Parkhi, and Andrew Zisserman. ``Vggface2: A dataset for recognising faces across pose and age.'' In \textit{IEEE International Conference on Automatic Face \& Gesture Recognition.} 2018.
 \bibitem[2]{simonyan2014very} Karen Simonyan and Andrew Zisserman. ``Very deep convolutional networks for large-scale image recognition'' \textit{arXiv preprint arXiv:1409.1556.} 2014.
  \bibitem[3]{he2016deep} He, Kaiming, Xiangyu Zhang, Shaoqing Ren, and Jian Sun. ``Deep residual learning for image recognition.'' In \textit{IEEE Conference on Computer Vision and Pattern Recognition.} 2016.
 \bibitem[4]{hu2018squeeze} Hu, Jie, Li Shen, and Gang Sun. ``Squeeze-and-excitation networks.'' In \textit{IEEE Conference on Computer Vision and Pattern Recognition.} 2018.
\end{thebibliography}

\end{comment}
